---
title: "NICHD Decoding Maternal Morbidity Data Challenge: predicting post-partum depression and assessing the socioeconomic impact"
author: 
  name: "Dr. Martin G. Frasch"
  affiliation: "Health Stream Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    theme: paper
    number_sections: yes
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message=TRUE, include=TRUE, warnings = FALSE, dev="png",dpi=300)

knitr::opts_chunk$set(echo = TRUE)
# be careful with cache setting: set to FALSE if you want to rerun all code before knitting
# more info: https://bookdown.org/yihui/rmarkdown-cookbook/cache.html

# syntax to time execution of a cell in R | added as time_it snippet | invoke with shift+TAB
ptm<-proc.time()
# insert my code
proc.time()-ptm
```

```{r, include=FALSE}
local({r <- getOption("repos"); r["CRAN"] <- "http://cran.us.r-project.org"; options(repos = r)})
if (!"R.utils" %in% rownames(installed.packages())) install.packages("R.utils")
setwd("/home/bizon/Documents/mu2b/")
options(echo=TRUE)
TEST_ROOT_DIR <- ".."
```

# Read the data

```{r load general packages for handling big data}
library(tidyverse)
library(data.table)
library(dplyr)
library(readxl)
library(caret)
library(magrittr)
```

```{r read the raw data, cache=TRUE}
ptm<-proc.time()
f<-"/home/bizon/Documents/nuMoM2b_Dataset_NICHD_Data_Challenge.csv"
mu_data_df <- fread(f)
proc.time()-ptm
```

## Set up easy in-the-app viewing of the accompanying coding and info spreadsheets

I want to be able to look up quickly and select the correct predictor and response variables directly inside this R Studio environment. Eventually, this could be extended to allow variable selection to pass them on to "predictors" and "response" variables used in the various machine learning (ML) models below.

For now, we are simply manually selecting the clinically/medically meaningful predictors/response variables, check them for artifacts, clean up those artifacts as appropriate and proceed to ML modeling to test our hypotheses.

```{r read the explanatory files to handle everything in one place, cache=TRUE}

ptm<-proc.time()
f1_info<-"/home/bizon/Documents/mu2b/nuMoM2b_Dataset_Information.xlsx"
f2_code<-"/home/bizon/Documents/mu2b/nuMoM2b_Codebook_NICHD_Data_Challenge.xlsx"
mu_info_df <- read_excel(f1_info)
mu_code_df <- read_excel(f2_code)
proc.time()-ptm
```

Now that we loaded our Information and Code spreadsheets, we can use DT library to view and search in them easily.

*Note: it is best to open the output tables in a new window.*

```{r view code and info tables, cache=TRUE}
library(DT)
datatable(mu_code_df, filter = 'top', options = list(pageLength = 10, autoWidth = TRUE))
datatable(mu_info_df, filter = 'top', options = list(pageLength = 10, autoWidth = TRUE))
```

# Preprocessing

[[I need to remove NA and non-sense values from]{.ul}]{.smallcaps} [CMAE04a4c[**.**]{.smallcaps}]{.ul}

*Note: this step can be repeated as needed in the preprocessing pipeline for the desired response variable*

```{r remove NA rows for CMAE04a4c, cache=TRUE}
# select "CMAE04a4c" | if NA, remove that row 

# remove all rows where the column CMAJ01 has NA
clean_CMAE04a4c_df<-mu_data_df[!is.na(mu_data_df$CMAE04a4c), ] 

# The above won't work on an h2o object, but works on a regular dataframe, so reading as df first and then, once preprocessing is done, moving on to h2o
```

```{r preprocessing_results_view_with_dplyr1, echo=TRUE, paged.print=TRUE, cache=TRUE}

glimpse(clean_CMAE04a4c_df$CMAE04a4c)
count(clean_CMAE04a4c_df, CMAE04a4c)
nrow(clean_CMAE04a4c_df[mu_data_df$CMAE04a4c])
sum(is.na(clean_CMAE04a4c_df$CMAE04a4c))
```

Only ten cases where PTSD postpartum CMAE04a4c=1. Not enough data for the ML approach at this time.

Exploring CMAE04a1c (postpartum depression).

```{r remove NA rows for CMAE04a1c, cache=TRUE}
# select "CMAE04a1c" | if NA, remove that row 

# remove all rows where the column CMAJ01 has NA
clean_CMAE04a1c_df<-mu_data_df[!is.na(mu_data_df$CMAE04a1c), ] 

# The above won't work on an h2o object, but works on a regular dataframe, so reading as df first and then, once preprocessing is done, moving on to h2o
```

```{r preprocessing_results_view_with_dplyr2, echo=TRUE, paged.print=TRUE, cache=TRUE}

glimpse(clean_CMAE04a1c_df$CMAE04a1c)
count(clean_CMAE04a1c_df, CMAE04a1c)
nrow(clean_CMAE04a1c_df[mu_data_df$CMAE04a1c])
sum(is.na(clean_CMAE04a1c_df$CMAE04a1c))
```

There are n=338 positive outcomes, so will pursue this in the following with the ML approaches as done for the re-hospitalization outcome before.

```{r load h2o}
library(h2o)
h2o.init()
```

```{r read_into_h2o, cache=TRUE}
ptm<-proc.time()
mu_data <- as.h2o(clean_CMAE04a1c_df)
proc.time()-ptm
```

```{r set_predictors_response, cache=TRUE}

mu_data[,"CMAE04a1c"]<-as.factor(mu_data[,"CMAE04a1c"]) 

# predictors "participant", "demographics" dataset (column A) includes the relevant features
# concludes the variables from "demographics" dataset; model as is and test later by adding participant features from the "CMA" set

predictors<-c("CRace",
              "Race",
              "eRace",
              "eHispanic",
              "BMI",
              "BMI_Cat",
              "Education",
              "GravCat",
              "SmokeCat1",
              "SmokeCat2",
              "SmokeCat3",
              "Ins_Govt",
              "Ins_Mil",
              "Ins_Comm",
              "Ins_Pers",
              "Ins_Othr",
              "PctFedPoverty",
              "poverty",
              "V1AD02g",                    #adding psych predictors here
              "CMAE04a1b",
              "CMAE04a4a", 
              "CMAE04a4b"
              ) #makes a list of predicting variables

response<-"CMAE04a1c"
```

# ML modeling

## Train the DRF model

(first, let's do a "quick&dirty" way = no hold-out dataset to get some sense of the data from ML standpoint)

```{r first_drf_model, cache=TRUE}

# user note: if you re-start h2o JVM, you must re-import the dataset and predictors/response vars
test2_mother_mu_data_RFmodel<-h2o.randomForest(x=predictors,y=response,training_frame = mu_data, nfolds=10,seed = 1234)
```

```{r model performance, cache=TRUE}

test2_mother_mu_data_RFmodel
```

    AUC:  0.9346045
    AUCPR:  0.6157573

That was a quick test.

Now I need to create a hold-out dataset first; repeat the above then as test:validation.

## Split the data 0.8 for training:validation

```{r data_split, cache=TRUE}
mother2_mu_data_split <- h2o.splitFrame(mu_data, ratios=0.8, seed = 1)
```

```{r assign_train_validation, cache=TRUE}
train=mother2_mu_data_split[[1]]
valid=mother2_mu_data_split[[2]]
```

## Train the DRF model on the training dataset = 0.8

```{r train_DRF_on_test_dataset, cache=TRUE}
ptm<-proc.time()
mother2_mu_data_RFmodel<-h2o.randomForest(x=predictors,y=response,training_frame = train, nfolds=10,seed = 1234)
proc.time()-ptm

```

## Predict using the DRF model on the testing dataset =0.2

> Yields AUC:  0.9288383 and AUCPR:  0.6338817

```{r performance of the _real_ DRF model, message=TRUE, paged.print=TRUE, cache=TRUE}
ptm<-proc.time()

mother2_mu_data_predict<-h2o.predict(object=mother2_mu_data_RFmodel, newdata=valid)

mother2_mu_data_RFmodel

proc.time()-ptm
```

### AUROCpr for predicting re-hospitalization based on patient demographics alone

```{r plot AUROCpr, echo=TRUE, cache=TRUE}

mod=mother2_mu_data_RFmodel

perf <- h2o.performance(mod,valid)

metrics <- as.data.frame(h2o.metric(perf))

metrics

metrics %>%
  ggplot(aes(recall,precision)) + 
  geom_line() +
  theme_minimal()

metrics %>%
  ggplot(aes(precision, accuracy)) + 
  geom_line() +
  theme_minimal()

```

### Explain the model

```{r model_performance_on_test_dataset, cache=TRUE}

ptm<-proc.time()

# toggle progress bar if desired:
# h2o.show_progress() 

exp <-h2o.explain(object=mother2_mu_data_RFmodel, newdata=valid)
print(exp)

proc.time()-ptm
```

### Statistics summary

```{r function to see the results, cache=TRUE}

results_df <- function(h2o_model) {
  h2o_model@model$cross_validation_metrics_summary %>% 
    as.data.frame() %>% 
    select(-mean, -sd) %>% 
    t() %>% 
    as.data.frame() %>% 
    mutate_all(as.character) %>% 
    mutate_all(as.numeric) -> k
  
  k %>% 
    select(Accuracy = accuracy,
           AUC = auc,
           Precision = precision,
           Specificity = specificity,
           Recall = recall,
           Logloss = logloss) %>% 
  return()
}
```

```{r producing the statistics summary, cache=TRUE}

# Using function
results_df(mod) -> outcome

# Outcome 
outcome %>% 
  gather(Metrics, Values) %>% 
  ggplot(aes(Metrics, Values, fill = Metrics, color = Metrics)) +
  geom_boxplot(alpha = 0.3, show.legend = FALSE) + 
  facet_wrap(~ Metrics, scales = "free") + 
  labs(title = "Performance of our ML model using H2o package ",
       caption = "Data Source: NICHD Decoding Maternal Morbidity Data Challenge\nCreated by Martin Frasch (further credit to https://bit.ly/3BpPqcb)") +
  theme_minimal()
```

```{r Tabular statistics summary, cache=TRUE}

# Statistics summary
outcome %>% 
  gather(Metrics, Values) %>% 
  group_by(Metrics) %>% 
  summarise_each(funs(mean, median, min, max, sd, n())) %>% 
  mutate_if(is.numeric, function(x) {round(100*x, 2)}) %>%
  knitr::kable(col.names = c("Criterion", "Mean", "Median", "Min", "Max", "SD", "N"))

```
## Building an interpretable decision tree model

```{r decision_tree_model, cache=TRUE}

maternal_dt_model<-h2o.gbm(x=predictors,y=response,training_frame = train, validation_frame = valid, balance_classes = TRUE, seed = 1234, nfolds=10)

# GBM hyperparamters
gbm_params = list(max_depth = seq(2, 10))

# Train and validate a cartesian grid of GBMs
gbm_grid = h2o.grid("gbm", x = predictors, y = response,
                    grid_id = "gbm_grid_1tree8",
                    training_frame = train,
                    validation_frame = valid,
                    balance_classes = TRUE,
                    ntrees = 1, min_rows = 1, sample_rate = 1, col_sample_rate = 1,
                    learn_rate = .01, seed = 1234, 
                    hyper_params = gbm_params)

gbm_gridperf = h2o.getGrid(grid_id = "gbm_grid_1tree8",
                           sort_by = "auc",
                           decreasing = TRUE)

```

```{r GBM_performance, cache=TRUE}

# what is the performance of this GBM?
maternal_dt_model
```

We obtain AUC:  0.9473067, AUCPR:  0.6510553

```{r GBM_grid_performance, cache=TRUE}
gbm_gridperf
```

Inflection point is at max_depth=5

```{r train GBM, cache=TRUE}

maternal_1_tree = h2o.gbm(x = predictors, y = response, 
                        training_frame = train, balance_classes = TRUE,
                        ntrees = 1, min_rows = 1, sample_rate = 1, col_sample_rate = 1,
                        max_depth = 5,
                  # use early stopping once the validation AUC doesn't improve by at least 0.01%
                  # for 5 consecutive scoring events
                        stopping_rounds = 3, stopping_tolerance = 0.01, 
                        stopping_metric = "AUC", 
                        seed = 1)
maternal_1_tree


```

    AUCPR:  0.882762

```{r maternal_tree, cache=TRUE}
maternal_Tree = h2o.getModelTree(model = maternal_1_tree, tree_number = 1)
```

```{r function_viz_tree, cache=TRUE}

# Visualizing H2O Trees

library(data.tree)

createDataTree <- function(h2oTree) {
  
  h2oTreeRoot = h2oTree@root_node
  
  dataTree = Node$new(h2oTreeRoot@split_feature)
  dataTree$type = 'split'
  
  addChildren(dataTree, h2oTreeRoot)
  
  return(dataTree)
}

addChildren <- function(dtree, node) {
  
  if(class(node)[1] != 'H2OSplitNode') return(TRUE)
  
  feature = node@split_feature
  id = node@id
  na_direction = node@na_direction
  
  if(is.na(node@threshold)) {
    leftEdgeLabel = printValues(node@left_levels, na_direction=='LEFT', 4)
    rightEdgeLabel = printValues(node@right_levels, na_direction=='RIGHT', 4)
  }else {
    leftEdgeLabel = paste("<", node@threshold, ifelse(na_direction=='LEFT',',NA',''))
    rightEdgeLabel = paste(">=", node@threshold, ifelse(na_direction=='RIGHT',',NA',''))
  }
  
  left_node = node@left_child
  right_node = node@right_child
  
  if(class(left_node)[[1]] == 'H2OLeafNode')
    leftLabel = paste("prediction:", left_node@prediction)
  else
    leftLabel = left_node@split_feature
  
  if(class(right_node)[[1]] == 'H2OLeafNode')
    rightLabel = paste("prediction:", right_node@prediction)
  else
    rightLabel = right_node@split_feature
  
  if(leftLabel == rightLabel) {
    leftLabel = paste(leftLabel, "(L)")
    rightLabel = paste(rightLabel, "(R)")
  }
  
  dtreeLeft = dtree$AddChild(leftLabel)
  dtreeLeft$edgeLabel = leftEdgeLabel
  dtreeLeft$type = ifelse(class(left_node)[1] == 'H2OSplitNode', 'split', 'leaf')
  
  dtreeRight = dtree$AddChild(rightLabel)
  dtreeRight$edgeLabel = rightEdgeLabel
  dtreeRight$type = ifelse(class(right_node)[1] == 'H2OSplitNode', 'split', 'leaf')
  
  addChildren(dtreeLeft, left_node)
  addChildren(dtreeRight, right_node)
  
  return(FALSE)
}

printValues <- function(values, is_na_direction, n=4) {
  l = length(values)
  
  if(l == 0)
    value_string = ifelse(is_na_direction, "NA", "")
  else
    value_string = paste0(paste0(values[1:min(n,l)], collapse = ', '),
                          ifelse(l > n, ",...", ""),
                          ifelse(is_na_direction, ", NA", ""))
  
  return(value_string)
}


```

This decision tree, also supplied as PDF, is meant to help build intuition about how the model.

```{r decision_tree_h2o, cache=TRUE}

library(DiagrammeR)

# customized DT for our H2O model

maternal_mu2DataTree = createDataTree(maternal_Tree)

GetEdgeLabel <- function(node) {return (node$edgeLabel)}
GetNodeShape <- function(node) {switch(node$type, 
                                       split = "diamond", leaf = "oval")}
GetFontName <- function(node) {switch(node$type, 
                                      split = 'Palatino-bold', 
                                      leaf = 'Palatino')}
SetEdgeStyle(maternal_mu2DataTree, fontname = 'Palatino-italic', 
             label = GetEdgeLabel, labelfloat = TRUE,
             fontsize = "26", fontcolor='royalblue4')
SetNodeStyle(maternal_mu2DataTree, fontname = GetFontName, shape = GetNodeShape, 
             fontsize = "26", fontcolor='royalblue4',
             height="0.75", width="1")

SetGraphStyle(maternal_mu2DataTree, rankdir = "LR", dpi=70.)

plot(maternal_mu2DataTree, output = "graph")

```

```{r generate h2o explanation of the DT model, cache=TRUE}
ptm<-proc.time()

exp_dt<-h2o.explain(maternal_dt_model,valid)

proc.time()-ptm
```

```{r show the explanation, cache=TRUE}

exp_dt
```

## Using **Naïve Bayes Classifier**

```{r Naïve Bayes Classifier, cache=TRUE}

# Build and train the model:
mo2b_nb <- h2o.naiveBayes(x = predictors,
                          y = response,
                          training_frame = train,
                          laplace = 0,
                          nfolds = 10,
                          seed = 1234)

# Eval performance:
perf <- h2o.performance(mo2b_nb)

# Generate the predictions on a test set (if necessary):
pred <- h2o.predict(mo2b_nb, newdata = valid)
```

```{r bayes_performance, cache=TRUE}
perf
```
NB model achieves AUC:  0.9259393 and AUCPR:  0.5121705
```{r explain naive bayes model, cache=TRUE}

# best viewed in a new window or see, please, the PDF included with the submission
exp_nb <- h2o.explain(mo2b_nb,valid)
```

Note the highly variable partial importance of the different socio-demographic characteristics

```{r vuz_nb_performance, cache=TRUE}

exp_nb
```

## Extending the ML toolbox: Using h2o AutoML mode to find an objectively best performing model

Compare [here](http://h2o-release.s3.amazonaws.com/h2o/rel-yates/1/docs-website/h2o-docs/automl.html#code-examples). The findings are to be interpreted with caution at this stage. Once we obtain a larger external dataset for validation, with a more balanced case distribution, this will become more useful and allow building an inference engine that could be deployed for use. I am presenting this code therefore as a reference for future work.

Nevertheless, it is evident that an optimization even at this stage results in a classification prediction performance of AUROC = 0.9937349. This result can vary depending on the run.

Note please, this code runs for about 70 min on a well-equipped deep learning workstation.

### Train in AutoML mode

```{r auto_ml, cache=TRUE}

ptm<-proc.time()

maternal_aml <- h2o.automl(x=predictors,y=response,training_frame = train, max_models = 20, seed = 1)

maternal_lb <- maternal_aml@leaderboard

#print(maternal_lb, n = nrow(maternal_lb)) #Print all rows instead of default 6 rows

proc.time()-ptm
```

### Validate the combined leader aml model

```{r validate_aml, cache=TRUE}
ptm<-proc.time()

maternal_perf_valid <- h2o.performance(maternal_aml@leader,newdata=valid,xval=FALSE,valid=TRUE)

pred <- h2o.predict(maternal_aml@leader, valid)
```

```{r}
h2o.auc(maternal_aml@leader, cache=TRUE)
```

[`Explain the model`](https://docs.h2o.ai/h2o/latest-stable/h2o-docs/explain.html)

We observe no specificity because the dataset is unbalanced such that by luck of draw (when the dataset is split 80:20) we get no true positives.

```{r explain_best_model, cache=TRUE}
ptm<-proc.time()
exp <-h2o.explain(maternal_aml@leader, valid)
proc.time()-ptm
```

```{r, cache=TRUE}
print(exp)

```

```{r}
sessionInfo()
```
