---
title: "NICHD Decoding Maternal Morbidity Data Challenge: prediction of rehospitalization"
author: 
  name: "Dr. Martin G. Frasch"
  affiliation: "Health Stream Analytics"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
    theme: paper
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# be careful with cache setting: set to FALSE if you want to rerun all code before knitting

# syntax to time execution of a cell in R | added as time_it snippet | invoke with shift+TAB
ptm<-proc.time()
# insert my code
proc.time()-ptm
```

```{r, include=FALSE}
local({r <- getOption("repos"); r["CRAN"] <- "http://cran.us.r-project.org"; options(repos = r)})
if (!"R.utils" %in% rownames(installed.packages())) install.packages("R.utils")
setwd("/home/bizon/Documents/mu2b/")
options(echo=TRUE)
TEST_ROOT_DIR <- ".."
```

# Read the data

```{r load general packages for handling big data}
library(tidyverse)
library(data.table)
library(dplyr)
library(readxl)
library(caret)
library(magrittr)
```

```{r read the raw data, cache=TRUE}
ptm<-proc.time()
f<-"/home/bizon/Documents/nuMoM2b_Dataset_NICHD_Data_Challenge.csv"
mu_data_df <- fread(f)
proc.time()-ptm
```

## Set up easy in-the-app viewing of the accompanying coding and info spreadsheets

I want to be able to look up quickly and select the correct predictor and response variables directly inside this R Studio environment. Eventually, this could be extended to allow variable selection to pass them on to "predictors" and "response" variables used in the various machine learning (ML) models below.

For now, we are simply manually selecting the clinically/medically meaningful predictors/response variables, check them for artifacts, clean up those artifacts as appropriate and proceed to ML modeling to test our hypotheses.

```{r read the explanatory files to handle everything in one place, cache=TRUE}

ptm<-proc.time()
f1_info<-"/home/bizon/Documents/mu2b/nuMoM2b_Dataset_Information.xlsx"
f2_code<-"/home/bizon/Documents/mu2b/nuMoM2b_Codebook_NICHD_Data_Challenge.xlsx"
mu_info_df <- read_excel(f1_info)
mu_code_df <- read_excel(f2_code)
proc.time()-ptm
```

Now that we loaded our Information and Code spreadsheets, we can use DT library to view and search in them easily.

*Note: it is best to open the output tables in a new window.*

```{r view code and info tables, cache=TRUE}
library(DT)
datatable(mu_code_df, filter = 'top', options = list(pageLength = 10, autoWidth = TRUE))
datatable(mu_info_df, filter = 'top', options = list(pageLength = 10, autoWidth = TRUE))
```

# Preprocessing

[[I need to remove NA values from **CMAJ01.**]{.ul}]{.smallcaps}

*Note: this step can be repeated as needed in the preprocessing pipeline for the desired response variable*

```{r remove NA rows for CMAJ01, cache=TRUE}
# select "CMAJ01" | if NA, remove that row 

# remove all rows where the column CMAJ01 has NA
clean_CMAJ01_df<-mu_data_df[!is.na(mu_data_df$CMAJ01), ] 

# The above won't work on an h2o object, but works on a regular dataframe, so reading as df first and then, once preprocessing is done, moving on to h2o
```

```{r preprocessing_results_view_with_dplyr, echo=TRUE, paged.print=TRUE, cache=TRUE}
#View(clean_CMAJ01_df$CMAJ01) #super fast Excel-like view of the data

glimpse(clean_CMAJ01_df$CMAJ01)
count(clean_CMAJ01_df, CMAJ01)
nrow(clean_CMAJ01_df[mu_data_df$CMAJ01])
sum(is.na(clean_CMAJ01_df$CMAJ01))
```

Confirming that we removed all NAs for CMAJ01:

    glimpse: int [1:8774] 2 2 2 2 2 2 2 2 2 2 ...
    count: [1] 9289
    sum is.na: [1] 0

+-----------------------------------------------------------------------------------------------------+
| Raw data cleaned with regard to NA values in CMAJ01, my response variable. Using this for modeling. |
+-----------------------------------------------------------------------------------------------------+

+===============================================================================+ +-------------------------------------------------------------------------------+

+--------+--------+
| CMAJ01 | n      |
|        |        |
| \<int> | \<int> |
+=======:+=======:+
| 1      | 154    |
+--------+--------+
| 2      | 8620   |
+--------+--------+

# Load h2o for Machine Learning

```{r load h2o}
library(h2o)
h2o.init()
```

Reading in the preprocessed dataset.

```{r read_into_h2o, cache=TRUE}
ptm<-proc.time()
mu_data <- as.h2o(clean_CMAJ01_df)
proc.time()-ptm

```

## Set predictors and response; set response as a factor

-   maternal outcome: mortality (CMAE14): don't have enough data to build a binary outcome prediction model \| skip;

-   First, exploring the impact of demographics on readmission rate (CMAJ01) as a measure of morbidity:

```{r set_predictors_response, cache=TRUE}

mu_data[,"CMAJ01"]<-as.factor(mu_data[,"CMAJ01"]) 

# predictors "participant", "demographics" dataset (column A) includes the relevant features
# concludes the variables from "demographics" dataset; model as is and test later by adding participant features from the "CMA" set

predictors<-c("CRace",
              "Race",
              "eRace",
              "eHispanic",
              "BMI",
              "BMI_Cat",
              "Education",
              "GravCat",
              "SmokeCat1",
              "SmokeCat2",
              "SmokeCat3",
              "Ins_Govt",
              "Ins_Mil",
              "Ins_Comm",
              "Ins_Pers",
              "Ins_Othr",
              "PctFedPoverty",
              "poverty"                        
              ) #makes a list of predicting variables

response<-"CMAJ01"
```

```{r view the dataset, cache=TRUE}
mu_data
```

```{r number of rows for CMAJ01, cache=TRUE}
h2o.nrow(mu_data["CMAJ01"])
h2o.group_by(data=mu_data, by="CMAJ01", nrow("CMAJ01"))
```

In the raw dataset, we had 8786 cases of "CMAE14" are coded as 2 = not died, there are no cases to study this outcome.

In the raw dataset, we had for "CMAJ01", we have a slightly better situation: 1, yes = 154 cases of readmission, 2, No, 8620 cases.

Needed to handle NaN and 4th row. This is best done on the R data frame which is what we did above in Preprocessing step.

```{r}
head(mu_data, cache=TRUE)
```

```{r h2o_aggregator_estimator}
# agg <- h2o.aggregator(training_frame = mu_data,
#                       target_num_exemplars = 9000,
#                       rel_tol_num_exemplars = 0.5,
#                       categorical_encoding = "Eigen")
# new_df<-h2o.aggregated_frame(agg)
# new_df
```

Aggregator destroys this dataframe. Skip this step.

# Train the DRF model

(first, let's do a "quick&dirty" way = no hold-out dataset to get some sense of the data from ML standpoint)

```{r first_drf_model, cache=TRUE}

test_mother_mu_data_RFmodel<-h2o.randomForest(x=predictors,y=response,training_frame = mu_data, nfolds=10,seed = 1234)
```

## Show the model's performance

    AUCpr:  0.9851791 for prediction of hospital readmission from demographics

```{r model performance, echo=TRUE, message=TRUE, warning=FALSE, cache=TRUE}

test_mother_mu_data_RFmodel
```

That was a quick test.

Now I need to create a hold-out dataset first; repeat the above then as test:validation.

See [here](https://docs.h2o.ai/h2o/latest-stable/h2o-r/docs/reference/index.html), [here](https://docs.h2o.ai/h2o/latest-stable/h2o-r/docs/reference/h2o.splitFrame.html) and [here](http://h2o-release.s3.amazonaws.com/h2o/master/3552/docs-website/h2o-docs/datamunge/splitdatasets.html)

## Split the data 0.8 for training:validation

```{r data_split, cache=TRUE}
mother_mu_data_split <- h2o.splitFrame(mu_data, ratios=0.8, seed = 1)
```

```{r assign_train_validation, cache=TRUE}
train=mother_mu_data_split[[1]]
valid=mother_mu_data_split[[2]]
```

## Train the DRF model on the training dataset = 0.8

```{r train_DRF_on_test_dataset, cache=TRUE}
ptm<-proc.time()
mother_mu_data_RFmodel<-h2o.randomForest(x=predictors,y=response,training_frame = train, nfolds=10,seed = 1234)
proc.time()-ptm

```

## Predict using the DRF model on the testing dataset =0.2

> Yields very good AUCpr=0.984

```{r performance of the _real_ DRF model, message=TRUE, paged.print=TRUE, cache=TRUE}
ptm<-proc.time()

mother_mu_data_predict<-h2o.predict(object=mother_mu_data_RFmodel, newdata=valid)

mother_mu_data_RFmodel

proc.time()-ptm
```

### AUROCpr for predicting re-hospitalization based on patient demographics alone

```{r plot AUROCpr, echo=TRUE, cache=TRUE}

mod=mother_mu_data_RFmodel

perf <- h2o.performance(mod,valid)

metrics <- as.data.frame(h2o.metric(perf))

metrics %>%
  ggplot(aes(recall,precision)) + 
  geom_line() +
  theme_minimal()

```

### Explain the model

```{r model_performance_on_test_dataset, cache=TRUE}

ptm<-proc.time()

# toggle progress bar if desired:
# h2o.show_progress() 

exp <-h2o.explain(object=mother_mu_data_RFmodel, newdata=valid)
print(exp)

proc.time()-ptm
```

### Statistics summary

```{r function to see the results, cache=TRUE}

results_df <- function(h2o_model) {
  h2o_model@model$cross_validation_metrics_summary %>% 
    as.data.frame() %>% 
    select(-mean, -sd) %>% 
    t() %>% 
    as.data.frame() %>% 
    mutate_all(as.character) %>% 
    mutate_all(as.numeric) -> k
  
  k %>% 
    select(Accuracy = accuracy,
           prAUC = pr_auc,
           Precision = precision,
           Specificity = specificity,
           Recall = recall,
           Logloss = logloss) %>% 
  return()
}
```

```{r producing the statistics summary, cache=TRUE}

# Using function
results_df(mod) -> outcome

# Outcome 
outcome %>% 
  gather(Metrics, Values) %>% 
  ggplot(aes(Metrics, Values, fill = Metrics, color = Metrics)) +
  geom_boxplot(alpha = 0.3, show.legend = FALSE) + 
  facet_wrap(~ Metrics, scales = "free") + 
  labs(title = "Performance of our ML model using H2o package ",
       caption = "Data Source: NICHD Decoding Maternal Morbidity Data Challenge\nCreated by Martin Frasch (further credit to https://bit.ly/3BpPqcb)") +
  theme_minimal()
```

```{r Tabular statistics summary, cache=TRUE}

# Statistics summary
outcome %>% 
  gather(Metrics, Values) %>% 
  group_by(Metrics) %>% 
  summarise_each(funs(mean, median, min, max, sd, n())) %>% 
  mutate_if(is.numeric, function(x) {round(100*x, 2)}) %>%
  knitr::kable(col.names = c("Criterion", "Mean", "Median", "Min", "Max", "SD", "N"))

```

## Building an interpretable decision tree model

> Using the `balance_classes` term available in this model allows to better control for the unbalanced data we are dealing with here, because the re-hospitalization rate is \~2%

[`Source`](http://rstudio-pubs-static.s3.amazonaws.com/463653_b50579f05ae246a9bfa4251ef9aae26b.html#plotting-decision-tree-with-data.table)

```{r decision_tree_model, cache=TRUE}

maternal_dt_model<-h2o.gbm(x=predictors,y=response,training_frame = train, validation_frame = valid, balance_classes = TRUE, seed = 1234, nfolds=10)

# GBM hyperparamters
gbm_params = list(max_depth = seq(2, 10))

# Train and validate a cartesian grid of GBMs
gbm_grid = h2o.grid("gbm", x = predictors, y = response,
                    grid_id = "gbm_grid_1tree8",
                    training_frame = train,
                    validation_frame = valid,
                    balance_classes = TRUE,
                    ntrees = 1, min_rows = 1, sample_rate = 1, col_sample_rate = 1,
                    learn_rate = .01, seed = 1234, 
                    hyper_params = gbm_params)

gbm_gridperf = h2o.getGrid(grid_id = "gbm_grid_1tree8",
                           sort_by = "auc",
                           decreasing = TRUE)

```

```{r GBM_performance, cache=TRUE}

# what is the performance of this GBM?
maternal_dt_model
```

We obtain prAUC=0.98

```{r GBM_grid_performance, cache=TRUE}
gbm_gridperf
```

Inflection point is at max_depth=8

```{r train GBM, cache=TRUE}

maternal_1_tree = h2o.gbm(x = predictors, y = response, 
                        training_frame = train, balance_classes = TRUE,
                        ntrees = 1, min_rows = 1, sample_rate = 1, col_sample_rate = 1,
                        max_depth = 8,
                  # use early stopping once the validation AUC doesn't improve by at least 0.01%
                  # for 5 consecutive scoring events
                        stopping_rounds = 3, stopping_tolerance = 0.01, 
                        stopping_metric = "AUC", 
                        seed = 1)
maternal_1_tree


```

    AUCPR:  0.882762

```{r maternal_tree, cache=TRUE}
maternal_Tree = h2o.getModelTree(model = maternal_1_tree, tree_number = 1)
```

```{r function_viz_tree, cache=TRUE}

# Visualizing H2O Trees

library(data.tree)

createDataTree <- function(h2oTree) {
  
  h2oTreeRoot = h2oTree@root_node
  
  dataTree = Node$new(h2oTreeRoot@split_feature)
  dataTree$type = 'split'
  
  addChildren(dataTree, h2oTreeRoot)
  
  return(dataTree)
}

addChildren <- function(dtree, node) {
  
  if(class(node)[1] != 'H2OSplitNode') return(TRUE)
  
  feature = node@split_feature
  id = node@id
  na_direction = node@na_direction
  
  if(is.na(node@threshold)) {
    leftEdgeLabel = printValues(node@left_levels, na_direction=='LEFT', 4)
    rightEdgeLabel = printValues(node@right_levels, na_direction=='RIGHT', 4)
  }else {
    leftEdgeLabel = paste("<", node@threshold, ifelse(na_direction=='LEFT',',NA',''))
    rightEdgeLabel = paste(">=", node@threshold, ifelse(na_direction=='RIGHT',',NA',''))
  }
  
  left_node = node@left_child
  right_node = node@right_child
  
  if(class(left_node)[[1]] == 'H2OLeafNode')
    leftLabel = paste("prediction:", left_node@prediction)
  else
    leftLabel = left_node@split_feature
  
  if(class(right_node)[[1]] == 'H2OLeafNode')
    rightLabel = paste("prediction:", right_node@prediction)
  else
    rightLabel = right_node@split_feature
  
  if(leftLabel == rightLabel) {
    leftLabel = paste(leftLabel, "(L)")
    rightLabel = paste(rightLabel, "(R)")
  }
  
  dtreeLeft = dtree$AddChild(leftLabel)
  dtreeLeft$edgeLabel = leftEdgeLabel
  dtreeLeft$type = ifelse(class(left_node)[1] == 'H2OSplitNode', 'split', 'leaf')
  
  dtreeRight = dtree$AddChild(rightLabel)
  dtreeRight$edgeLabel = rightEdgeLabel
  dtreeRight$type = ifelse(class(right_node)[1] == 'H2OSplitNode', 'split', 'leaf')
  
  addChildren(dtreeLeft, left_node)
  addChildren(dtreeRight, right_node)
  
  return(FALSE)
}

printValues <- function(values, is_na_direction, n=4) {
  l = length(values)
  
  if(l == 0)
    value_string = ifelse(is_na_direction, "NA", "")
  else
    value_string = paste0(paste0(values[1:min(n,l)], collapse = ', '),
                          ifelse(l > n, ",...", ""),
                          ifelse(is_na_direction, ", NA", ""))
  
  return(value_string)
}


```

This decision tree, also supplied as PDF, is meant to help build intuition about how the model.

```{r decision_tree_h2o, cache=TRUE}

library(DiagrammeR)

# customized DT for our H2O model

maternal_mu2DataTree = createDataTree(maternal_Tree)

GetEdgeLabel <- function(node) {return (node$edgeLabel)}
GetNodeShape <- function(node) {switch(node$type, 
                                       split = "diamond", leaf = "oval")}
GetFontName <- function(node) {switch(node$type, 
                                      split = 'Palatino-bold', 
                                      leaf = 'Palatino')}
SetEdgeStyle(maternal_mu2DataTree, fontname = 'Palatino-italic', 
             label = GetEdgeLabel, labelfloat = TRUE,
             fontsize = "26", fontcolor='royalblue4')
SetNodeStyle(maternal_mu2DataTree, fontname = GetFontName, shape = GetNodeShape, 
             fontsize = "26", fontcolor='royalblue4',
             height="0.75", width="1")

SetGraphStyle(maternal_mu2DataTree, rankdir = "LR", dpi=70.)

plot(maternal_mu2DataTree, output = "graph")

```

```{r generate h2o explanation of the DT model, cache=TRUE}
ptm<-proc.time()

exp_dt<-h2o.explain(maternal_dt_model,valid)

proc.time()-ptm
```

```{r show the explanation, cache=TRUE}

exp_dt
```

## Using **Naïve Bayes Classifier**

```{r Naïve Bayes Classifier, cache=TRUE}

# Build and train the model:
mo2b_nb <- h2o.naiveBayes(x = predictors,
                          y = response,
                          training_frame = train,
                          laplace = 0,
                          nfolds = 10,
                          seed = 1234)

# Eval performance:
perf <- h2o.performance(mo2b_nb)

# Generate the predictions on a test set (if necessary):
pred <- h2o.predict(mo2b_nb, newdata = valid)
```

```{r bayes_performance, cache=TRUE}
perf
```

    AUC:  0.6178245
    AUCPR:  0.9877748

```{r explain naive bayes model, cache=TRUE}

# best viewed in a new window or see, please, the PDF included with the submission
exp_nb <- h2o.explain(mo2b_nb,valid)
```

Note the highly variable partial importance of the different socio-demographic characteristics

```{r viz_nb_performance, cache=TRUE}

exp_nb
```

## Extending the ML toolbox: Using h2o AutoML mode to find an objectively best performing model

Compare [here](http://h2o-release.s3.amazonaws.com/h2o/rel-yates/1/docs-website/h2o-docs/automl.html#code-examples). The findings are to be interpreted with caution at this stage. Once we obtain a larger external dataset for validation, with a more balanced case distribution, this will become more useful and allow building an inference engine that could be deployed for use. I am presenting this code therefore as a reference for future work.

Nevertheless, it is evident that an optimization even at this stage results in a marked (\~60%) improvement of classification prediction performance, up to AUC 76%. This result can vary depending on the run.

Note please, this code runs for \~2 hours on a well-equipped deep learning workstation.

### Train in AutoML mode

```{r auto_ml, cache=TRUE}

ptm<-proc.time()

maternal_aml <- h2o.automl(x=predictors,y=response,training_frame = train, max_models = 20, seed = 1)

maternal_lb <- maternal_aml@leaderboard

#print(maternal_lb, n = nrow(maternal_lb)) #Print all rows instead of default 6 rows

proc.time()-ptm
```

### Validate the combined leader aml model

```{r validate_aml, cache=TRUE}
ptm<-proc.time()

maternal_perf_valid <- h2o.performance(maternal_aml@leader,newdata=valid,xval=FALSE,valid=TRUE)

pred <- h2o.predict(maternal_aml@leader, valid)
```

```{r aml_auc, cache=TRUE}
h2o.auc(maternal_aml@leader)
```

```{r aml_metrics, cache=TRUE}
# Using function
results_df(maternal_aml@leader) -> outcome

# Outcome 
outcome %>% 
  gather(Metrics, Values) %>% 
  ggplot(aes(Metrics, Values, fill = Metrics, color = Metrics)) +
  geom_boxplot(alpha = 0.3, show.legend = FALSE) + 
  facet_wrap(~ Metrics, scales = "free") + 
  labs(title = "Performance of the best AutoML model using H2o package ",
       caption = "Data Source: NICHD Decoding Maternal Morbidity Data Challenge\nCreated by Martin Frasch (further credit to https://bit.ly/3BpPqcb)") +
  theme_minimal()
```

[`Explain the model`](https://docs.h2o.ai/h2o/latest-stable/h2o-docs/explain.html)

We observe no specificity because the dataset is unbalanced such that by luck of draw (when the dataset is split 80:20) we get no true positives.

```{r explain_best_model, cache=TRUE}
ptm<-proc.time()
exp <-h2o.explain(maternal_aml@leader, valid)
proc.time()-ptm
```

```{r explain, cache=TRUE}
print(exp)
```

# GPU accelerated ML solutions

Test run from the [demo](https://docs.h2o.ai/h2o/latest-stable/h2o-r/docs/reference/h2o.randomForest.html) site. Details on setup [here](https://cran.r-project.org/web/packages/h2o4gpu/vignettes/getting_started.html).

### Testing h2o GPU installation

Requires Python h2o4gpu module. Details of installation [here](https://github.com/h2oai/h2o4gpu).

-   The R library uses reticulate to point to the correct Python version (3.6)

-   [Vignettes](https://github.com/h2oai/h2o4gpu/blob/master/src/interface_r/vignettes/getting_started.Rmd) \| Python [demos](https://github.com/h2oai/h2o4gpu/tree/master/examples/py/demos)

```{r a relevant example of GPU-accelerated ML, cache=TRUE}
library(h2o4gpu)
library(reticulate)  # only needed if using a virtual Python environment or multiple Python versions
use_python("/home/bizon/anaconda3/bin/python3.6", required = TRUE)
#use_virtualenv("/home/bizon/h2o_ml")  # set this to the path of your venv if any is used

# Setup dataset
x <- iris[1:4]
y <- as.integer(iris$Species) - 1

# Initialize and train the classifier
model <- h2o4gpu.random_forest_classifier() %>% fit(x, y)

#  Other ML approaches are also available such as GBM, regression models, unsupervised learning

# Make predictions
predictions <- model %>% predict(x)
```

### Deploying GPU-accelerated ML on our dataset

I am leaving this for future implementations. This is simply to point out the possibility to leverage GPU-equipped deep learning workstations to speed up the model building time on a dataset of this size.

=> See "postpartum depression" notebook for additional models.

```{r, cache=TRUE}
sessionInfo()
```
